Explore bias sensitivity of Vieira-Silva et al (2019) QMP analysis

### 2022-02-09 notebook/_posts/2022-01-20-vieirasilva2019quan/vieirasilva2019quan.Rmd
Revise and render vieirasilva2019quan analysis
Analysis of bias sensitivity of Spearman correlations on QMP profiles
from vieirasilva2019quan.

---
title: "Explore Vieira-Silva et al (2019) QMP profiles"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - vieirasilva2019quan
date: 2022-01-20
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

@@

@@

```{r libraries}
library(here)
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

@@

@@

Note, there is an extra value in the BMI column of Table S1 (Excel spreadsheet).

```{r}
otu <- path(data_path, 'qmp-matrix-genus.tsv') %>%
  read.table() %>%
  as('matrix') %>%
  otu_table(taxa_are_rows = FALSE)
sam <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S1', range = 'A2:L174', na = "NA")
original_column_names <- names(sam)
ps <- phyloseq(
  otu, 
  sam %>% janitor::clean_names() %>% sample_data
)
stopifnot(setequal(sample_names(ps), sample_names(otu)))
sam <- ps %>% sample_data %>% as_tibble %>% glimpse
```

### Taxonomy

how can I get taxonomy and/or phylogeny for the genus identifiers?

First, create a file with the RDP 16 taxonomy from the DADA2 training file

```
zcat rdp_train_set_16.fa.gz | grep '^>' | sort | uniq -u | sed -e 's/^>//' -e 's/;$//' - > rdp_train_set_16_headers.txt
```

```{r}
cns <- c('domain', 'phylum', 'class', 'order', 'family', 'genus')
cts <- 'cccccc'
# x <- path('~/data/16s-taxonomy-databases/dada2/rdp-trainset-16',
#   'rdp_train_set_16_headers.txt') %>%
#   read_delim(delim = ';', col_names = cns, col_types = cts)
x <- path('~/data/16s-taxonomy-databases/dada2/rdp-trainset-16',
  'rdp_train_set_16_headers.txt') %>%
  read_lines %>%
  enframe %>%
  select(-name) %>%
  separate(value, into = cns, sep = ';', fill = 'right') %>%
  glimpse
```

```{r}
genera <- taxa_names(ps) %>% 
  str_subset('unclassified', negate = TRUE)
genera %in% x$genus
setdiff(genera, x$genus)

unclassified_taxa <- taxa_names(ps) %>% 
  str_subset('unclassified') %>%
  str_extract('(?<=unclassified_).+')
unclassified_taxa %in% x$family
unclassified_taxa %in% x$order
unclassified_taxa %in% x$phylum
unclassified_taxa %in% x$domain
```

try again

```{r}
rnks <- c('domain', 'phylum', 'class', 'order', 'family', 'genus')
x <- path('~/data/16s-taxonomy-databases/rdp/', 
  'RDPClassifier_16S_trainsetNo16_rawtrainingdata',
  'trainset16_022016_headers.txt') %>%
  read_tsv(col_names = c('id', 'path'))
y <- x %>%
  separate(path, c('root', rnks), sep = ';', fill = 'right')

```

ok, giving up..

hmm - check to see if can get from the rrndb tables?
https://rrndb.umms.med.umich.edu/static/download/

# Explore

I think that the QMP profiles are such that the sum within a sample should equal the reported cell counts per gram in the sample data; let's check.

```{r}
all.equal(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
```

This small difference may be down to rounding error, since it seems that the entires in the QMP matrix are rounded to integers.

```{r}
otu %>% c %>% {all.equal(., round(.))}
sam$average_faecal_cell_count_cells_g %>% {all.equal(., round(.))}
```

There are `r nsamples(ps)`, with the following disease breakdown:

@@

Can we recover counts?
We can't recover the raw counts without using the CNV information.
But setting that aside for now, let's confirm that we understand the normalization that was used -

```{r}
x <- sam %>%
  add_column(sample_sum = sample_sums(ps))
```
```{r}
x %>% pull(sample_sum) %>% summary
x %>% transmute(sample_sum / average_faecal_cell_count_cells_g) %>% pull(1) %>% summary
```

(we already knew this - the samples sum up to the cell density).

# Reproduce paper analysis

Let's try to reproduce the results from the paper.

@@

First, let's look for taxa associated with faecal calprotectin concentrations

```{r}
x <- ps %>% as_tibble
```

```{r}
y <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  as_tibble %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman_test = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman', # exact = FALSE, continuity = TRUE,
        )),
    spearman_test = map(spearman_test, broom::tidy)
  ) %>%
  select(-data) %>%
  unnest(spearman_test) %>%
  mutate(p.bh = p.adjust(p.value, method = 'BH'))
```

```{r}
y %>% arrange(p.bh) %>% print(n=20)
```

```{r}
study_res <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S6', na = "NA")
study_res %>%
  select(Genera, starts_with('Calprotectin') & ends_with('QMP'))
y %>% arrange(p.bh) %>% filter(!str_detect(.otu, 'unclassified')) %>%
  select(.otu, estimate, p.value, p.bh) %>%
  print(n=15)
```

These estimates match, but the p values (and their ranking) differs.

figure this out; then try to recapitulate using rank regression.

Note, a decrease FDR-corrected p values can be expected if the FDR correction was done _after_ discarding the OTUs without genus classifications.

can check the code at https://github.com/raeslab/QMP/

```{r}
study_res %>%
  select(Genera, starts_with('Calprotectin') & ends_with('QMP')) %>%
  arrange(4) %>%
  print(n=15)
y %>% 
  arrange(p.bh) %>% 
  filter(!str_detect(.otu, 'unclassified')) %>%
  mutate(
    p.bh = p.adjust(p.value, method = 'BH'),
    across(starts_with('p.'), round, digits = 4)
  ) %>%
  select(.otu, estimate, p.value, p.bh) %>%
  print(n=15)
```

ahhh - ok their's is sorted differently, which makes sense since it has many tests in one table. and the p values need to be rounded to 4 digits. then it seems like we're good.

next:

- also do for moisture, and perhaps RMP, to check these results as well
- confirm I can get similar result with linear model on ranks
<!--  -->

visually inspect

```{r}
top_hits <- y %>% 
  filter(!str_detect(.otu, 'unclassified')) %>%
  slice_min(p.bh, n = 9)
x <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  prune_taxa(top_hits$.otu, .) %>%
  as_tibble

x %>%
  mutate(across(.otu, factor, levels = top_hits$.otu)) %>%
  with_groups(.otu, mutate,
    across(c(faecal_calprotectin_mg_g, .abundance), rank),
  ) %>%
  # mutate(across(c(faecal_calprotectin_mg_g, .abundance), rank, ties.method = 'average')) %>%
  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
  facet_wrap(~.otu) +
  geom_point() +
  stat_smooth(method = 'lm')
```

It seems like zeros or below-detection-limit values are largely responsible for the observed results.

```{r}
x %>%
  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~.otu) +
  geom_point()
```

### spearman versus linear modeling

NOTE: The results differ when I first omit the samples with missing calprotectin measurements, or not.
perhaps because the ranks of the abundances differ?

```{r}
z <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  as_tibble %>%
  with_groups(.sample, mutate, proportion = close_elts(.abundance)) %>%
  with_groups(.otu, mutate, 
    across(c(.abundance, proportion, faecal_calprotectin_mg_g), 
      rank, ties.method = 'average',
      .names = '{.col}_rank')
  ) %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman', na.action = na.omit)),
    lm = map(data, 
      ~lm(data = .x, .abundance_rank ~ faecal_calprotectin_mg_g_rank,
        na.action = na.omit)),
  ) %>%
  select(-data) %>%
  pivot_longer(c(spearman, lm), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
z1 <- z %>% filter(test == 'spearman' | term != '(Intercept)')
```

NOTE: The p values are nearly identical as long as I first drop the samples with missing data.
But not if I don't.

```{r}
z1 %>%
  select(.otu, test, p.value) %>%
  pivot_wider(names_from = test, values_from = p.value) %>%
  ggplot(aes(spearman, lm)) +
  geom_abline(color = 'darkgrey') +
  scale_x_log10() +
  scale_y_log10() +
  geom_point()
```

```{r}
z1 %>%
  select(.otu, test, estimate) %>%
  pivot_wider(names_from = test, values_from = estimate) %>%
  ggplot(aes(spearman, lm)) +
  geom_abline(color = 'darkgrey') +
  # scale_x_log10() +
  # scale_y_log10() +
  geom_point()
```

The estimates are related but differ - I suspect this has to do with ties in the abundance (0s) and in the calprotectin (min values / detection limit). Perhaps the problematic taxa are very low prevalence?

### try all tests

```{r}
y <- ps %>%
  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
  as_tibble %>%
  with_groups(.sample, mutate, proportion = close_elts(.abundance)) %>%
  with_groups(.otu, nest) %>%
  mutate(
    calprotectin_qmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman')),
    calprotectin_rmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ proportion + faecal_calprotectin_mg_g, 
        method = 'spearman')),
    moisture_qmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + moisture_content_percent, 
        method = 'spearman')),
    moisture_rmp_spearman = map(data, 
      ~cor.test(data = .x,
        ~ proportion + moisture_content_percent, 
        method = 'spearman')),
  ) %>%
  select(-data) %>%
  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
```

TODO: check that samples with NAs are handled in the spearman tests

```{r}
y.w <- y %>%
  select(.otu, test, estimate, p.bh) %>%
  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
  select(.otu, sort(tidyselect::peek_vars())) %>%
  mutate(across(where(is.numeric), round, 4)) %>%
  arrange(calprotectin_qmp_spearman_p.bh)
```

Note, I'm not sure what the RMP values used in the original study were.
From the manuscript, it's not obvious if the unclassified taxa were removed before the proportions were computed, or if copy number correction was used.

### Check if patterns are driven by presence/absence

```{r}
y <- ps %>%
  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
  as_tibble %>%
  with_groups(.sample, mutate, 
    proportion = close_elts(.abundance),
    present = (.abundance > 0) * 1
  ) %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman')),
    spearman_pa = map(data, 
      ~cor.test(data = .x,
        ~ present + faecal_calprotectin_mg_g, 
        method = 'spearman')),
  ) %>%
  select(-data) %>%
  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
```

```{r}
y.w <- y %>%
  select(.otu, test, estimate, p.bh) %>%
  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
  select(.otu, sort(tidyselect::peek_vars())) %>%
  mutate(across(where(is.numeric), round, 4)) %>%
  arrange(spearman_p.bh)
```

```{r}
y.w %>%
  ggplot(aes(spearman_estimate, spearman_pa_estimate)) +
  geom_point()
```

For some taxa (like Veillonella, EscherichiaShigella) the two estimates are very similar;for others like Streptococcus they are quite different.
I suspect that simulated bias will have a larger impact on Streptococcus.

@@

```{r, eval = F}
# Perturb by bias vector - treat ps as actual
ps.per <- perturb(ps, bias, norm = 'keep')
# Calibrate by bias vector - treat ps as observed
ps.cal <- calibrate(ps, bias, norm = 'keep')
```

Maintain the sample sums, but perturb the relative abundances.

```{r}
n_taxa <- ntaxa(ps)
# Standard deviation for the distribution of log efficiencies
sigma.log_eff <- 3
# helper function to draw a single random bias vector
rbias <- function() exp(rnorm(n_taxa, 0, sigma.log_eff))
# helper function to calibrate the phyloseq object by the bias vector
my_cal <- function(bias, ps) {
  ps %>%
    calibrate(bias, norm = 'keep')
}

set.seed(42)
# Number of replicates
R <- 3e2
# data frame with R random bias vectors
reps <- seq(R) %>%
  set_names %>%
  map(~rbias()) %>%
  enframe("id", "bias") %>%
  mutate(
    actual = map(bias, my_cal, ps = ps)
  )
```

```{r}
my_stat <- function(ps, otu) {
  data <- ps %>%
    prune_taxa(otu, .) %>%
    as_tibble
  spearman_test <- cor.test(data = data,
    ~ .abundance + faecal_calprotectin_mg_g, 
    method = 'spearman', exact = FALSE
  )
  spearman_test %>% broom::tidy()
}

#> focal_otu <- 'Veillonella'
original <- my_stat(ps, focal_otu)
reps_veill <- reps %>%
  mutate(
    .otu = 'Veillonella',
    fit = map2(actual, .otu, my_stat)
  ) %>%
  unnest(fit)
reps_strep <- reps %>%
  mutate(
    .otu = 'Streptococcus',
    fit = map2(actual, .otu, my_stat)
  ) %>%
  unnest(fit)
```

```{r}
plot_bias_sens <- function(reps) {
  est_dist <- ggplot(reps, aes(estimate)) +
    geom_histogram() +
    geom_vline(xintercept = original$estimate, color = "darkred") +
    geom_vline(xintercept = 0, color = "grey") +
    labs(title = "Distribution of point estimates", x = "Estimate", y = "Count")
  pval_dist <- ggplot(reps, aes(p.value)) +
    geom_histogram() +
    geom_vline(xintercept = original$p.value, color = "darkred") +
    geom_vline(xintercept = 0.05, color = "grey") +
    scale_x_log10() +
    labs(title = "Distribution of p-values", x = "p-value", y = "Count")
  est_dist / pval_dist
}
# ci_reps <- ggplot(bias_reps, aes(x = estimate, y = as.integer(id))) +
#   geom_point(size = 0.8) +
#   geom_errorbarh(aes(xmin = conf.low, xmax = conf.high)) +
#   geom_point(data = original, aes(y = R*1.05), 
#     size = 0.8, color = "darkred") +
#   geom_errorbarh(data = original, 
#     aes(xmin = conf.low, xmax = conf.high, y = R*1.05),
#     color = "darkred") +
#   geom_vline(xintercept = 0, color = "grey") +
#   theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
#   labs(title = "Replicate 95% CIs", x = "Estimate", 
#     y = "Replicate")
# plot_grid(
#     plot_grid(est_dist, pval_dist, ncol = 1, labels = c("A", "B")),
#     ci_reps,
#     labels = c("", "C")
#     )
```

```{r}
plot_bias_sens(reps_veill)
```

```{r}
plot_bias_sens(reps_strep)
```

Example where the bias sensitivity analysis shows little effect of varying the hypothesized bias.

We can probably understand this finding because the original result is driven by the presence/absence pattern, and we have treated zeros as absolute, so that presence/absence is maintained.
If we relaxed this assumption, then I suspect that we would see much more variation in the estimates.
Much of this variation, however, would come from relaxing our assumption about the zeros, rather than relaxing our assumption that there is no bias.

Challenge with this data is that we don't have the read counts, so don't know how to replace zeros with resonable values.

## saved plotting stuff

Mean efficiency distribution:

```{r}
x <- reps %>%
  slice(1) %>%
  pull(actual) %>%
  .[[1]] %>%
  sample_data %>%
  as_tibble

x %>%
  ggplot(aes(x = mean_efficiency, fill = faecal_calprotectin_mg_g > 50)) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x %>%
  ggplot(aes(x = faecal_calprotectin_mg_g, y = mean_efficiency, color = diagnosis)) +
  scale_y_log10() +
  scale_x_log10() +
  geom_point() +
  stat_smooth(aes(group = 1), method = 'lm')
```

Now, redo the analysis with and without bias

what assumptions to make about zeros?

>>>>

---
title: "Explore bias sensitivity of Vieira-Silva et al (2019) QMP analysis"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - ref:vieirasilva2019quan
  - bias sensitivity
  - absolute abundance
date: 2022-01-20
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

@@

This document reanalyzes the 'Quantitative Microbiome Profiles' of @vieirasilva2019quan.

@@

```{r libraries}
library(here)
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(ggridges)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

@@

The genus-level 'Quantitative Microbiome Profiles' from @vieirasilva2019quan are available for download at http://raeslab.org/software/QMP2/, and the sample metadata is available in the supplemental content at the journal site.

@@

I'll read in the metadata and QMP profiles into a phyloseq object.
The metadata is Table S1 of the Excel spreadsheet.
There is an extra value in the BMI column of Table S1 that will be dropped in the import step below.

```{r}
otu <- path(data_path, 'qmp-matrix-genus.tsv') %>%
  read.table() %>%
  as('matrix') %>%
  otu_table(taxa_are_rows = FALSE)
sam <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S1', range = 'A2:L174', na = "NA")
original_column_names <- names(sam)
ps <- phyloseq(
  otu, 
  sam %>% janitor::clean_names() %>% sample_data
) %>%
  rename_sample_data(
   faecal_calprotectin = faecal_calprotectin_mg_g,
   moisture_content = moisture_content_percent
  )
stopifnot(setequal(sample_names(ps), sample_names(otu)))
sam <- ps %>% sample_data %>% as_tibble %>% glimpse
```

# Explore data

There are `r nsamples(ps)`, with the following counts by disease status,

@@

The QMP profiles are such that the sum within a sample should equal the reported cell counts per gram in the sample data; let's check.

```{r}
all.equal(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
cor(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
```

This small difference may be down to rounding error, since apparently the entires in the QMP matrix have been rounded to integers:

```{r}
otu %>% c %>% {all.equal(., round(.))}
```

It would be nice to have read counts; however, I don't think it's possible to obtain read counts from the available information here.
Perhaps it would be possible to reverse the QMP calculation if we had the copy-number-variation information.

Since we lack read-count information, we will ignore the sequencing-count noise, which was also done in the original paper.

Our primary covariates of interest are Faecal calprotectin and moisture content.

```{r}
sam %>% select(faecal_calprotectin, moisture_content) %>% summary
```

```{r}
sam %>%
  ggplot(aes(x = faecal_calprotectin)) +
  scale_x_log10() +
  geom_histogram()
```

Note that a large fraction of calprotectin measurements are precisely at 30, suggesting this is a value assigned to samples that were below the detection limit of the assay.
Also, the values vary over multiple orders of magnitude.

```{r}
sam %>%
  ggplot(aes(x = moisture_content)) +
  geom_histogram()
```

In contrast, the moisture varies relatively continuously

# Reproduce QMP Spearman analysis

Let's try to reproduce the results from the paper.
We'll focus on the Spearman correlations of the genera absolute abundances with faecal calprotectin and stool moisture content.
The relevant manuscript text is

@@

A summary table with the Spearman results are Table S6 of the supplementary data Excel sheet,

```{r}
study_res <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S6', na = "NA")
```

```{r}
qmp_spearman <- ps %>%
  prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
  as_tibble %>%
  with_groups(.otu, nest) %>%
  mutate(
    qmp_calprotectin_spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin, 
        method = 'spearman', exact = FALSE, continuity = TRUE)),
    qmp_moisture_spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + moisture_content, 
        method = 'spearman', exact = FALSE, continuity = TRUE)),
  ) %>%
  select(-data) %>%
  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
```

Compare Calprotectin results,

```{r}
study_res %>%
  select(Genera, starts_with('Calprotectin') & ends_with('QMP')) %>%
  arrange(`Calprotectin  Spearman.P QMP`) %>%
  slice_head(n=12)
qmp_spearman %>%
  filter(test == 'qmp_calprotectin_spearman') %>%
  arrange(p.value) %>% 
  select(.otu, estimate, p.value, p.bh) %>%
  slice_head(n=12)
```

The results are similar but not identical.

Compare moisture results,

```{r}
study_res %>%
  select(Genera, starts_with('Moisture') & ends_with('QMP')) %>%
  arrange(`Moisture Spearman.P QMP`) %>%
  slice_head(n=12)
qmp_spearman %>%
  filter(test == 'qmp_moisture_spearman') %>%
  arrange(p.value) %>% 
  select(.otu, estimate, p.value, p.bh) %>%
  slice_head(n=12)
```

Here the estimate and p-values seem identical up to rounding, but the FDR-corrected p values differ somewhat.

Note, I filtered the OTUs that lack genus classifications prior to doing the testing, which I suspect is what was done in the original study.
If these OTUs are kept, the FDR-corrected p-values are larger, due to the larger number of tests.
But the study results have smaller p-values and only list the OTUs with genus names in the table.

## Visually inspect variation for top hits

### Calprotectin

```{r}
top_hits <- qmp_spearman %>% 
  filter(test == 'qmp_calprotectin_spearman') %>%
  slice_min(p.bh, n = 9)
x <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin)) %>%
  prune_taxa(top_hits$.otu, .) %>%
  as_tibble %>%
  mutate(across(.otu, factor, levels = top_hits$.otu))
```

To visualize the spearman correlation, plot and perform a linear regression on the rank-transformed values,

```{r, fig.dim = c(8,8)}
x %>%
  with_groups(.otu, mutate,
    across(c(faecal_calprotectin, .abundance), rank),
  ) %>%
  ggplot(aes(faecal_calprotectin, .abundance)) +
  facet_wrap(~.otu) +
  coord_fixed() +
  geom_point() +
  stat_smooth(method = 'lm') +
  labs(y = 'rank(abundance)', x = 'rank(faecal calprotectin)')
```

Many of the top hits have relatively low pravalence, and it appears that the correlation is being driven by differential prevalence in the two groups of samples where calprotectin has its min value, or is above its min value.
For these taxa, we might expect bias to have less impact on the associations.

### Moisture

```{r}
top_hits <- qmp_spearman %>% 
  filter(test == 'qmp_moisture_spearman') %>%
  slice_min(p.bh, n = 9)
x <- ps %>%
  filter_sample_data(!is.na(moisture_content)) %>%
  prune_taxa(top_hits$.otu, .) %>%
  as_tibble %>%
  mutate(across(.otu, factor, levels = top_hits$.otu))
```

```{r, fig.dim = c(8,8)}
x %>%
  with_groups(.otu, mutate,
    across(c(moisture_content, .abundance), rank),
  ) %>%
  ggplot(aes(moisture_content, .abundance)) +
  facet_wrap(~.otu) +
  coord_fixed() +
  geom_point() +
  stat_smooth(method = 'lm') +
  labs(y = 'rank(abundance)', x = 'rank(moisture content)')
```

Here we instead see that top taxa as being more prevalent.

## Check if hits are driven by differential prevalence

One way we can do this is to also do Spearman tests after converting abundances to presence/absence (0 if absent, 1 if present).
If the original result is driven by prevalence, then the estimates should be similar to when the QMP abundance was used.

```{r}
qmp_presence <- ps %>%
  prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
  as_tibble %>%
  mutate(
    presence = (.abundance > 0) * 1
  ) %>%
  with_groups(.otu, nest) %>%
  mutate(
    qmp_calprotectin = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin, 
        method = 'spearman', exact = FALSE, continuity = TRUE)),
    presence_calprotectin = map(data, 
      ~cor.test(data = .x,
        ~ presence + faecal_calprotectin, 
        method = 'spearman', exact = FALSE, continuity = TRUE)),
    # cond.qmp_calprotectin = map(data,
    #   ~cor.test(data = .x %>% filter(.abundance > 0),
    #     ~ .abundance + faecal_calprotectin, 
    #     method = 'spearman', exact = FALSE, continuity = TRUE)),
    qmp_moisture = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + moisture_content, 
        method = 'spearman', exact = FALSE, continuity = TRUE)),
    presence_moisture = map(data, 
      ~cor.test(data = .x,
        ~ presence + moisture_content, 
        method = 'spearman', exact = FALSE, continuity = TRUE)),
    # cond.qmp_moisture = map(data,
    #   ~cor.test(data = .x %>% filter(.abundance > 0),
    #     ~ presence + moisture_content, 
    #     method = 'spearman', exact = FALSE, continuity = TRUE)),
  ) %>%
  select(-data) %>%
  pivot_longer(contains(c('qmp', 'presence')), 
    names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
```

```{r}
qmp_presence_wide <- qmp_presence %>%
  select(.otu, test, estimate, p.value, p.bh) %>%
  separate(test, into = c('response', 'covariate'), sep = '_') %>%
  pivot_wider(names_from = response, values_from = c(estimate, p.value, p.bh), 
    names_glue = '{response}_{.value}', names_sort = TRUE) %>%
  select(.otu, covariate, starts_with('qmp'), starts_with('presence'))
#> select(.otu, sort(tidyselect::peek_vars()))
```

```{r}
p1 <- qmp_presence_wide %>%
  ggplot(aes(presence_estimate, qmp_estimate)) +
  facet_wrap(~covariate) +
  coord_fixed() +
  geom_point()
p1
```

```{r}
p2 <- qmp_presence_wide %>%
  ggplot(aes(presence_p.value, qmp_p.value)) +
  facet_wrap(~covariate) +
  coord_fixed() +
  scale_x_log10() +
  scale_y_log10() +
  geom_point()
p2
```

Overall, can see that most of the QMP correlations are highly similar to the presence correlations.

```{r, echo = FALSE}
print_table <- function(x, covariate) {
  x %>%
    filter(covariate == !!covariate) %>%
    select(-covariate) %>%
    rename(genus = .otu) %>%
    slice_min(qmp_p.value, n = 10) %>%
    mutate(across(where(is.numeric), signif, 2)) %>%
    knitr::kable()
}
```

**Faecal calprotectin**

```{r, echo = FALSE}
print_table(qmp_presence_wide, 'calprotectin')
```

**Moisture content**

```{r, echo = FALSE}
print_table(qmp_presence_wide, 'moisture')
```

Consider the calprotectin results.
For some taxa (like Veillonella, EscherichiaShigella) the two estimates are very similar;for others like Streptococcus they are quite different.

@@

Hypothesis: Correlations that are driven by differential prevalence will be less sensitive to bias than correlations that are not.

For the purposes of this analysis, zero observations will be treated as true zeros, which I expect to reduce the apparent impact of bias perturbation or calibration on the DA results relative to if we assumed zeros as indicating small positive abundances.

## Analysis

In order to perform a bias sensitivity analysis on the QMP profiles, we will calibrate (inverse-perturb) the relative abundances by the sampled bias vector, while keeping the total sum for the same constant, so that the total abundance of the same remains given by the fecal cell count.

```{r}
# Returns a matrix with taxa as columns
sample_bias <- function(n, gsd, vcv, taxa_as = 'cols') {
  stopifnot(isSymmetric(vcv))
  stopifnot(n > 1)
  n_taxa <- nrow(vcv)
  gsd_log <- log(gsd)
  bias <- MASS::mvrnorm(n, mu = rep(0, n_taxa), Sigma = vcv) %>%
    t %>%
    scale %>%
    {. * gsd_log} %>%
    exp
  if (taxa_as == 'cols')
    bias <- t(bias)
  bias
}

my_stat <- function(ps, otus) {
  ps %>%
    prune_taxa(otus, .) %>%
    as_tibble %>%
    with_groups(.otu, nest) %>%
    mutate(
      fit = map(data, ~cor.test(data = .x,
          ~ .abundance + faecal_calprotectin, 
          method = 'spearman', exact = FALSE, continuity = TRUE)),
      fit = map(fit, broom::tidy)
    ) %>%
    select(-data) %>%
    unnest(fit)
}
#> my_stat(ps, 'Veillonella')
```

```{r}
set.seed(42)
R <- 300
gsd <- 3
vcv <- diag(ntaxa(ps))
reps <- sample_bias(n = R, gsd = gsd, vcv = vcv) %>%
  split(., row(.)) %>%
  enframe(".idx", "bias")
```

```{r}
taxa_to_test <- c(
  #  low prevalence
  'Veillonella', 'EscherichiaShigella', 'Fusobacterium',
  # high prevalence
  'Streptococcus', 'Anaerostipes', 'Fusicatenibacter'
  )
reps_fit <- xfun::cache_rds({
  reps %>%
    mutate(
      fits = map(bias, 
        ~my_stat(
          ps = calibrate(ps, .x, norm = 'keep'),
          otus = taxa_to_test
        ))
    ) %>%
    unnest(fits)
}, dir = '_cache/', file = 'reps_fit', 
hash = list(reps, taxa_to_test))
```

```{r}
plot_bias_sens <- function(reps, original, .otu) {
  est_dist <- ggplot(reps, aes(estimate)) +
    geom_histogram() +
    geom_vline(xintercept = original$estimate, color = "darkred") +
    geom_vline(xintercept = 0, color = "grey") +
    labs(title = "Distribution of point estimates", x = "Estimate", y = "Count")
  pval_dist <- ggplot(reps, aes(p.value)) +
    geom_histogram() +
    geom_vline(xintercept = original$p.value, color = "darkred") +
    geom_vline(xintercept = 0.05, color = "grey") +
    scale_x_log10() +
    labs(title = "Distribution of p-values", x = "p-value", y = "Count")
  est_dist / pval_dist +
    plot_annotation(title = .otu)
}

original <- qmp_presence %>%
  filter(.otu %in% taxa_to_test, test == 'qmp_calprotectin') %>%
  with_groups(.otu, nest) %>%
  rename(original = data)

reps_plots <- reps_fit %>%
  with_groups(.otu, nest) %>%
  rename(reps = data) %>%
  left_join(original, by = ".otu") %>%
  mutate(.,
    plot = pmap(., plot_bias_sens)
  ) %>%
  select(.otu, plot) %>%
  deframe
```

## Top hits with low prevalence

```{r, echo = F}
fig.dim_sens <- c(6,6)
```

```{r, echo = FALSE, fig.dim = fig.dim_sens}
reps_plots$Veillonella
```

```{r, echo = FALSE, fig.dim = fig.dim_sens}
reps_plots$EscherichiaShigella
```

```{r, echo = FALSE, fig.dim = fig.dim_sens}
reps_plots$Fusobacterium
```

## Top hits with high prevalence

```{r, echo = FALSE, fig.dim = fig.dim_sens}
reps_plots$Streptococcus
```

```{r, echo = FALSE, fig.dim = fig.dim_sens}
reps_plots$Anaerostipes
```

```{r, echo = FALSE, fig.dim = fig.dim_sens}
reps_plots$Fusicatenibacter
```

## Summary

It isn't clear from these plots that there is a difference in bias sensitivity between the two sets.
